---
title: "01162019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Good practice to tell folks what packages to install up front for users 

```{r, warning = F, message = F}
library(dplyr)
library(tidyr)
```
to call a functino from a specific packate 'package_name::fucntino_name(...)'

# Data Cleaning
# 

```{r}
catch_df <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1",
                      method = "libcurl"),
                  stringsAsFactors = FALSE)

head(catch_df)
```

control shift m for %>% 

get ride of columns done wannt or select

* Remove  marginal sum and notes column
* move from wide to long format

```{r}

catch_long<- catch_df %>% 
  select(-All, -notesRegCode) %>% # negative calues remote columns
  gather(key = 'species', value = "catch", -Year, -Region)

head(catch_long)

```

Here we fix the misrepresentation of # of fish 


```{r}
 catch_cleaned <- catch_long %>% 
  rename(catch_thousands = catch) %>% #%>% 
  mutate(catch_thousands = ifelse(catch_thousands == "I", 1, catch_thousands)) %>%  # 1 means true 
  mutate(catch_thousands = as.integer(catch_thousands)) %>% # this added to make sure catch_thousands is not a cht but and int
 #mutate(catch = catch_thousands * 1000)
  mutate(catch = catch_thousands * 1000)

tail(catch_cleaned)
```
if see this error: Error in mutate_impl(.data, dots) : Evaluation error: non-numeric argument to binary operator.

then there is a letter in the catch value column ....yike
so comment out some things above then do a test to find the issue

```{r, eval = F, echo = F}
#test_catch <- as.integer(catch_cleaned$catch_thousands)

#i <- which(is.na(test_catch) == T) # to look for issue

#catch_cleaned[i, ] #to find issue first item is rown and second in columns in bracket

```

```{r}
catch_total <- catch_cleaned %>% 
  group_by(Region, Year) %>% 
  summarize(catch_region = sum(catch))

catch_total
```
# try to do average catch a year of
```{r}
catch_matt <- catch_cleaned %>% 
  group_by(species, Year) %>% 
  summarize(catch_mean = mean(catch),
            catch_sd = sd(catch),
              catch_n = n()) # total observatinos
          

catch_matt
```
# Split-Apply-Combine

Filter for Chinook 

```{r}
catch_chinook <- catch_cleaned %>% 
  filter(species == "Chinook" & Region == "SSE" & Year > 1990) %>% 
  arrange(-Year)
#  -Year desecingin
head(catch_chinook) 
```


#JOins


```{r}
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", method = "libcurl"),
                        stringsAsFactors = FALSE)
head(region_defs) # this lets you look at the a sample of the data.
```


# clean out column I do not want
```{r}
region_clean <- region_defs %>% 
  select(code, mgmtArea)

head(region_clean)
```
  
  #Join
```{r}
   catch_joined <- left_join(catch_cleaned, region_clean, 
                             by = c("Region" = "code"))
head(catch_joined)
test <- c("Region", "code")
                             
                             
```

#Spread

make a shorter dataframe that is wider....where coulmn are indivudal years
```{r}
catch_wide <- catch_cleaned %>% 
  filter(Year > 1990) %>% 
  select(-catch_thousands) %>% 
  spread(key = Year, value = catch)

head(catch_wide)
```

#Sepearte adn Unite

```{R}
dates_df <- data.frame(date = c("5/24/1930",
                                "5/25/1930",
                                "5/26/1930",
                                "5/27/1930",
                                "5/28/1930"),
                       stringsAsFactors = FALSE)

dates_df %>% 
  separate(date, c("month", "day", "year"), "/")
##   month day year
## 1     5  24 1930
## 2     5  25 1930
## 3     5  26 1930
## 4     5  27 1930
## 5     5  28 1930
```

```{r}
dates_sep <- dates_df %>% 
  separate(col = date, into = c("month", "day", "year"), by = "/", remove = F)

head(dates_sep)
```

#Unite

```{r}
dates_unite <- dates_sep %>% 
  unite(date_is0, year, month, day, sep = "-") # makes date look ISO standard

head(dates_unite)
```
# date
?as.PSCIXct
look up strip tie
formatsys.date %y%....